Cheminformatics: The Computer Science of Chemical Discovery

### 2011-11-09 rebuttal.tex

\comment{If this manuscript were to be published it would be
  completely rewriting the history of cheminformatics}
\response{While we agree with the reviewer that some key milestones in
  the history of cheminformatics were not included in the original
  article, we believe that the addition of the supplemental history
  addresses this problem. Furthermore, our original aim was \emph{not}
  to provide a complete historical overview of the field. Others have
  already provided such overviews. Our aim was to present a brief
  overview of \emph{topics} in cheminformatics, selected because we
  believed that they represent on going challenges for the field and
  thus amenable to coss-disciplinary efforts.}

\fbox{\textbf{Reviewer 3}}

\comment{page 1, column 1, line 24: the authors need to provide a
  reference for the assertion that cheminformatics is an older field,
  particularly given the comments later in this
  manuscript.}
\response{We have removed the assertion that cheminformatics is older
  than bioinformatics. Primarily, because there is no obvious
  reference that claims this. But alse because, depending on what one
  considers bioinformatics and cheminformatics, one field can be
  identified as being older or younger than the other. We feel that
  claims on age do not strengthen the paper and have thus removed it}

\comment{p1, c1, l29-37: I would dispute that until recently the
  cheminformatics techniques have been closely guarded
  secrets. J. Chem. Doc. was founded in 1961 and other techniques even
  older, such as the Wiener index (1947), Wiswesser Line Notation
  (1949), etc. Not to mention the actual foundations of what we call
  cheminformatics back to the atomistic theory of the 19th century.}
\response{We have the updated the text on page 1 to attenuate this
  statement, stressing on the fact that much of the data in
  cheminformatics is proprietary, whereas only some of the techniques (such
  as SMILES canonicalization) are. Regarding the statement that atomistic
  theory laid the foundations of cheminformatics - this is certainly
  true in a very broad sense. For that matter the atomistic theory
  laid the ground for computational chemistry in general. In the
  supplemental history document, we make a distinction between
  computational chemistry and cheminformatics}

\comment{p1, c1, l32: I think the phrase miracle molecule could be
  more usefully replaced with new drugs or new small molecule
  therapeutics.}
\response{The text has been updated to use the term ``therapeutic
  molecule''}

\comment{P4, c2, l41: the GDB-13 database contains 970 million
  molecules, which is nearly a billion, not a trillion.}
\response{The text has been updated to use the correct number}

\comment{P5, c1, l29: the normal phrase used to describe this concept
  is the similar property principle. The authors should also provide a
  reference.}
\response{This portion of the text has been restructured to refer to
  the similarity property principle and also include the relevant reference}

\end{document}

### 2011-11-09 paper.tex
Addressed more reviewer points

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, the
  data and techniques employed in the cheminformatics field have been
  closely guarded secrets of companies whose financial success
  depended on being the first to produce the new ``miracle molecule.''
  Only within the last decade---and as an effect of a change in
  mindsets, government-mandated policy changes and last but not least
  because of chemists volunteering their time for an Open Science
  ``movement''---have researchers gained access to freely available
  software packages and databases of tens of millions of
  chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding a cure for malaria and other ``neglected'' diseases.
\end{abstract}

@@

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a trillion chemical
structures~\cite{GDB}. Yet,
the current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates molecules up to 13 heavy atoms---given the possibly
exponential increase in search space with the number of atoms, it seems unlikely that the practically relevant chemical
space of about 30 heavy atoms will be exhaustively enumerated in the near future.

@@

The basis of predictive modeling in cheminformatics is that
\emph{chemical structure implies biological activity}. Thus, the goal
of any modeling approach is to capture and characterize the
correlations between structural features and the observed biological
activities. The goal is to describe cause-effect relationships with
the lowest possible risk, and also to describe the likelihood of error
(confidence) when using such models for decision making.  Figure~\ref{figure:applicability-domain} (in the Appendix) illustrates the use of substructure matching to produce a space of chemical structures that are likely to share similar biological activity; this is called the generation of an \emph{applicability domain}.

In many cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
focusing only on small molecule features, and therefore loose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.

>>>>

\maketitle
\begin{abstract}
  One of the most prominent success stories in all the sciences over
  the last decade has been the advance of bioinformatics: The
  interdisciplinary collaboration between computer scientists and
  molecular biologists led as one of its major accomplishments to the
  sequencing of the human genome, where researchers connected
  biological concepts to the theory of string algorithms. However,
  despite this great success, few computer scientists are familiar
  with a related discipline: cheminformatics, the use of computers to
  represent the structures of small molecules and analyze their
  properties.  Cheminformatics has wide applicability, from the
  discovery of novel drug candidates to the optimization of certain
  physicochemical properties of small molecules. Until recently, much
  of the the data and some of the techniques employed in the
  cheminformatics field have been closely guarded secrets of companies
  whose financial success depended on being the first to produce the
  new therapeutic molecules.  Only within the last decade---and as an
  effect of a change in mindsets, government-mandated policy changes
  and last but not least because of chemists volunteering their time
  for an Open Science ``movement''---have researchers gained access to
  freely available software packages and databases of tens of millions
  of chemicals. Academic chemists now confront a variety of unsolved
  algorithmic problems that could not have been tackled a decade ago,
  but whose solutions are critical to research ranging from
  determining the behavior of small molecules in biological pathways,
  to finding a cure for malaria and other ``neglected'' diseases.
\end{abstract}

@@

The underlying computational complexity and potential intractability
of molecular enumeration may not be a bottleneck after all, as
computational chemists have developed and successfully used
enumeration tools to generate large chemical libraries such as GDB-13,
comprising almost a billion chemical structures~\cite{GDB}. Yet, the
current enumeration software products do not generally produce
stereoisomers or tautomers; these require specific enumeration
procedures, which are still the subject of investigations by the
cheminformatics community. In addition, GDB-13 only enumerates
molecules up to 13 heavy atoms---given the possibly exponential
increase in search space with the number of atoms, it seems unlikely
that the practically relevant chemical space of about 30 heavy atoms
will be exhaustively enumerated in the near future.

@@

The basis of predictive modeling in cheminformatics is that the
biological activity of a molecule is a function of its chemical
structure. Together with the \emph{similarity property principle}
\cite{Johnson:1990ys} that states that structurally similar molecules
will exhibit similar properties, the goal of any modeling approach is
to capture and characterize the correlations between structural
features and the observed biological activities. Simultaneously, such
approaches must also describe the likelihood of error (confidence)
when using such models for decision making.  A variety of approaches
can be employed to assess the error in (or conversely, the reliability
of) a prediction, ranging from statistical approaches (XXX) to more
empirical approaches such as defining an \emph{applicability domain} -
the region of the input (i.e., chemical structures) that can be
reliably predicted, usually deliniated using some form of similarity
to the training set.

In many cases, the activity of a small molecule is due to its interaction with
a receptor. Traditional ``QSAR'' approaches ignore receptor interactions,
focusing only on small molecule features, and therefore lose valuable
information. As a result, techniques such as proteochemometric
modeling have been designed to take into account both ligand and
receptor structures.
